Modelling Interactive Computing Systems: Do We Have a Good Theory of What Computers Are?

### 2023-02-01 ART_Martin/Martin.tex
ToscanoMartin et al. article added. Minor style adjustments (sections, fancyhdr)

\documentclass[a4paper, 11pt, twoside]{article}

%Citation-related commands
%PP\usepackage{natbib}
%PP\setcitestyle{authoryear,open={(},close={)}} 
%PP \bibliographystyle{agsm}

\usepackage[utf8]{inputenc}
\usepackage{hyperref}
%PP \renewcommand\harvardyearleft{\unskip, }
%PP \renewcommand\harvardyearright[1]{.}
%PP \hypersetup{hidelinks}

\usepackage[top=3cm, bottom=3cm, left=3cm, right=3cm]{geometry}
%\usepackage{amsmath}
%\usepackage{amssymb}
%\usepackage{mathrsfs}
\usepackage{lmodern}
%\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{dsfont}
\usepackage{listings}
\usepackage[table]{xcolor}
%\usepackage{mathrsfs}
\usepackage{float}
\usepackage[justification = centering]{caption}
%\usepackage{amsmath}
\usepackage{graphicx}
%\usepackage{lmodern}
%\usepackage{bm}

%PP added
\usepackage[style=bath, citestyle=authoryear-comp, backend=biber]{biblatex} %uwaga: wymaga zmiany procesora bibliografii na biber
\addbibresource{Martin.bib}
%end of PP added

\title{Modelling Interactive Computing Systems: Do We Have a Good Theory of What Computers Are?}
%\date{}
\author{Alice Martin, Mathieu Magnaudet, Stéphane Conversy}

\begin{document}

\maketitle

Computers are increasingly interactive. They are no more transformational systems producing a final output after a finite execution. Instead, they continuously react in time to external events that modify the course of computing execution. While philosophers have been interested in conceptualizing computers for a long time, they seem to have paid little attention to the specificities of interactive computing. We propose to tackle this issue by surveying the literature in theoretical computer science, where one can find explicit proposals for a model of interactive computing. 
In that field, the formal modelling of interactive computing systems has been brought down to whether the new interaction models are reducible to Turing Machines. There are three areas where interaction models are framed. The comparison between TMs and interactive system models is at stake in all of them. These areas are namely some works on concurrency by Milner, on Reactive Turing Machines, and on interaction as a new computing paradigm. For each of the three identified models, we present its motivation, sum up its account for interaction and its legacy, and point out issues regarding the understanding of computers. The survey shows difficulties for epistemologists. The reason is that these analyses focus on the formal equivalence between interactive models of computation and classic ones. Such a project is different from addressing how a computing machine can be interactive: in other words, which mechanisms allow it. \\

Keywords: philosophy of computing, models of computation, interactive computing, computing mechanism, computational mechanistic explanation

@@

In the philosophy of computing, we are paying increased attention to a set of new features of computers. This set has led to the introduction of a new label for these computing machines: they are referred to as interactive computing machines \parencite{Dodig-Crnkovic2011, Goldin2006, Soare2013, VanLeeuwen2001, Wegner1997}. The set of new features can be captured in the following statement made in a 2011 paper by Gordana Dodig-Crnkovic (our highlights):

\begin{quote}
 Present day computers are very different from the early stand-alone calculators designed for mechanizing mathematical operations. They are largely \emph{used for communication in world-wide networks} and variety of information processing and knowledge management. Moreover, they play \emph{an important role in the control of physical processes and thus connect to the physical world}, especially in automation and robotics. [...] Computational processes are nowadays \emph{distributed, reactive, agent-based, and concurrent}. The main criterion of success of the computation is not its termination, but its \emph{response to the outside world}, its speed, generality and flexibility; adaptability, and tolerance to noise, error, faults, and damage. \parencite{Dodig-Crnkovic2011}  \end{quote}

Historically, the concept of interaction was introduced by a computer scientist, Milner, in the 1970s-1980s \parencite{Milner1975, Milner1982, Milner1993, Milner1999}. At first, an \emph{interactive computing system} was defined as a system where several threads execute instructions in parallel while being able to synchronize and communicate at certain moments of the execution. Since then, the characteristics of computer systems have continued to evolve, and by ``interactive'' we refer today to a broader set of properties that can be grouped as follows: the ability to continuously react in time to external events that modify computing execution. 
This class of computers deserves all our attention since they are ubiquitous. Every computer system today is designed to respond to external events in a predictable way and according to temporal constraints. In any case, what distinguishes this class of so-called interactive computing machines from the classical computer systems that preceded them is that they are no longer purely \emph{transformational systems}. A \emph{transformational system} is a classical computing device that, given a set of inputs, produces a final output after a finite execution. 
This evolution of computing complicates the answer to \textit{what a computer is}. The question is well-known in the philosophy of computing \parencite{Piccinini2008, Rapaport2018, Smith2002}. As already noted, many answers to the question distort it and answer the question of \textit{what a computation is}, immediately projecting the field of investigation into the theory of computability:

\begin{quote}
A fairly obvious, trivial, and almost-circular definition of ‘computer’ says that a computer is a machine that computes. The natural next question is: What does it mean to compute? But this shifts the burden of answering our question away from what computers are to the topic of what computation is. Many of the objections to various theories about computers are really objections to what counts as a computation. \parencite{Rapaport2018}. 
\end{quote}

This leaves us with the specific issue we want to address. We ask whether models of computation for interaction allow us to answer the question of what an actual (necessarily interactive) computer is. Current computers come in various forms and we chose in this paper to restrict our concerns to a delimited notion of \textit{interaction}, as defined in Human-Computer Interaction \parencite{Myers1994, Dearden1997, Beaudouin-Lafon2006,  Hornbaek2017, Basman2018}, and target a specific set of ubiquitous computing devices---those interacting with humans, e.g., through digital interfaces. We will not elaborate on \textit{analog computing} \parencite{Bielecki2019} and \textit{natural computation} \parencite{Dodig-Crnkovic2011, MacLennan2003}.

To tackle the issue of interactive computing devices, we propose here an approach that, to the best of our knowledge, has not been proposed so far: we want to examine the models of computation proposed in theoretical computer science to think about interactive computing systems. We offer a literature survey where one can find \textit{explicit theories of interaction} \footnote{We insist on our two criteria: \textit{explicit theories of interaction} in \textit{theoretical computer science}. We have in mind the fact that other communities e.g., the engineering community on reactive systems, are related to our topic but they have not conceptualized \textit{interaction} as such.}. We show that the formal modelling of interactive computing systems has been brought down to whether the new interaction models are reducible to Turing’s \emph{a-machines} \parencite{Turing1937}---we will refer to them as Turing Machines (TMs). 
Questioning the theoretical bounds of the Turing Machine in computer science when faced with the existence of interactive computing devices has been explored at least since Milner's work on communicating and mobile systems \parencite{Milner1993, Milner1999}. To the best of our knowledge, there are three areas where interaction models are framed as such. These areas are some works (i) on concurrency by Milner and his followers \parencite{Milner1999, Milner2006}, (ii) on Reactive Turing Machines \parencite{Andersen1997, Baeten2013, VanLeeuwen2001, VanLeeuwen2006}, and (iii) on interaction as a new non-algorithmic computing paradigm \parencite{Goldin2006, Wegner1997, Wegner2003}. For each of the three identified models, we:

@@

We then want to show how these approaches, which belong to a formal approach, cannot provide an answer to the question of what computers are, and for two reasons. On the one hand, these models of computation have focused their attention on whether interactive models are reducible to models of classical computation---par excellence, the Turing machine. Proving (or not) that an interactive property can be formalized as a computable property in the classical Turing’s sense does not answer the question of how an interactive property comes into existence and can be the object of execution. On the other hand, and this is a correlate, these models do not propose a basis for a mechanistic explanation of the very possibility of an interactive computing system. With only formal models of interactive computation, we might run the risk of not offering an adequate conceptualization of current computers. Therefore, we end up proposing to take the distinction seriously between \emph{models of computation} and \emph{mechanistic computational explanations}, as presented by Miłkowski \parencite*{Mikowski2011, Mikowski2014}.

\section{Milner introduces a distinction between interactional and computational behavior}
\label{Milner}

\subsection{Motivation}

Milner was the one who introduced the concept of interaction in computer science. He summarized his motivations in a famous Turing Award speech \parencite{Milner1993}. Milner was concerned with the logical foundations of computing inherited from Turing. He was preoccupied with the idea that computing practices had evolved since the birth of computing, notably in terms of architecture. He took seriously the possibility that the logical foundations dating back to the thirties may not match the growing challenges of his time and may require additional concepts.

Milner \parencite*{Milner2006} pointed out  that the logical foundations of computing offered by Turing \parencite*{Turing1937} were previous to the first physical computers and that computer science is grounded in \emph{logic} and \emph{engineering}. On the engineering side, computer science was inherited from von Neumann’s pioneering work \parencite{Godfrey1993, Aspray1990}. Only one thing could happen at once in an early von Neumann's computer. Nevertheless, there was more to computing than von Neumann's architecture \parencite{Backus1978, Milner2006}. A growing interest in dealing with concurrency in the sixties and seventies made sequential programming less warranted. Therefore, to Milner, the logical foundations of computing were to evolve. The main flaw of the early logical foundations was the reduction of computing processes to the concept of an algorithm, which tends to associate computing with mere calculation without taking concurrent activity into account. Because of the evolution of computing engineering practice, Milner questioned whether the logical grounds of computing should evolve as well. Milner's thesis can be put in a nutshell: ``this logical foundation has changed a lot since Turing but harks back to him. To be more precise: (i) Computing has grown into informatics---the science of interactive systems; (ii) Thesis: Turing's logical computing machines are matched by a logic of interaction'' \parencite{Milner2006}. Consequently, a theory and new language to express concurrent activity were required: ``we must find an elementary model which does for interaction what Turing's logical machines do for computation'' \parencite{Milner2006}.

The need to define a new computing theory is first displayed through the evolution of computing practice.
To sum up, Milner’s motivation and focus were the solving of concurrency issues in distributed systems, with the idea that the evolution of computing practices required new formal tools: ``Through the 1970s, I became convinced that a theory of concurrency and interaction requires a new conceptual framework, not just a refinement of what we find natural for sequential [algorithmic] computing'' \parencite{Milner1993}.

@@

Milner introduced the opposition between \textit{interactional} and \textit{computational} behaviour. Introducing the concept of interaction, Milner \parencite*{Milner1975, Milner1982, Milner1983} referred to concurrent message passing between agents. Milner's work coincided with Petri’s \parencite*{Petri1980} new model of concurrent processes, which generally intended to describe concurrency in information systems.\footnote{Concurrency theory emerged from Dick Karp's early work in the 1960s, grew with \parencite{Petri1980} and later work on transition systems \parencite{Nielsen1981, Glabbeek2004},
and has now developed into a mature theory of reactive systems \parencite{Harel1985} with diverse network models \parencites[for an overview, see][]{Lee1998}{Lee2006_modelsofcomputation}}
 To Milner, interaction is more \textit{expressive} than a TM, but it still describes an \textit{effective} procedure. Milner did not assert \textit{equivalence} between an interactive model and a TM, but he introduced the topic \parencite{Milner1999} and left it unanswered. 
Four main differences between old (computational) and new (interactional) computing are made striking by Milner. First, in Milner’s words, a Turing Machine prescribes a behaviour to be executed. In contrast, new computing requires the description of an information flow between several system components. Second, old computing is characterized by a \textit{hierarchical design} when current practice involves \textit{heterarchical} phenomena in the computing system. Third, in new computing, the designer cannot predict when agents will be triggered or the overall behaviour of the computing system. Fourth, the user is not merely looking for an end result in new computing practice. There is more than a mathematical function to evaluate, as it used to be in old computing. The user instead interacts with the system, and the look for an end result is replaced by continuing interaction. Having taken stock of the evolution of computing practice on the engineering side, Milner examines its consequences on the logic foundations of computing. The pi-calculus and his work on the equivalence with automata, known as bisimulation, achieved this reflection on interactive processes \parencite{Milner1993, Milner1999} with a formalism.

@@

Milner's work on interaction has become a founding block in automata theory and concurrency theory. It installed the notion of a transition system as the prime mathematical model to represent discrete behavior \parencite{Nielsen1981, Baldan2001, Glabbeek2004, Arbach2015}. It also showed that language equivalence was not the correct notion when comparing automata for interactive systems.
Instead, it should be replaced by a notion of behavioral equivalence or bisimilarity~\parencite{Milner1999}. The pi-calculus has inspired research to derive a language from it. The \textit{Pict} \parencite{Pierce2000} programming language is an example.

Milner's work is foundational and served as a reference for anyone after him, reflecting on the need for a new framework dedicated to new emerging computing practices. Milner insists on an essential reminder that we would like to consider. When modelling, the engineering practice matters and is to be articulated with the logical foundations of the model, should it involve elaborating a new framework. Famously, Wegner and Goldin acknowledge that Milner was the first to introduce the idea that classic models of computation were insufficient. They argue that Milner did not state clearly whether computation in concurrent communicating systems (CCS) and the pi-calculus were reducible to Turing machines and algorithms \parencite{Wegner2003}. If one goes and looks at Milner’s Turing Award Speech, it seems true that classical computation translates into an interactive calculus. However, it is not stated whether any formula in the pi-calculus can be expressed in a classical calculus like the lambda-calculus.

@@

Given the account of current computers that we are looking for, we see two limits in the lessons drawn from Milner.
First, we are looking for an explanation of the interactive computing phenomena at stake in a computer. Therefore, the relation between layers of abstraction, from the computational to the physical, is crucial. However, to Milner, the physical layer of the machine is not of much interest, and the calculus of CCS needs to abstract away from the physical. As Milner puts it, informatics is about virtual symbols: ``physical systems tend to have permanent physical links; they have fixed structure. But most systems in the informatic world are not physical; their links may be virtual or symbolic'' \parencite{Milner1999}. From our perspective, abstracting away from the physical world comes at some cost for an explanation. A complete explanation of a computing system can hardly be provided in details within a single understandable abstraction, since a computing sytems is extremely  multi-layered~\parencite{Nisan2005, Plato2018}. Therefore, an explanation of a computing system is necessarily a trade-off between understandability and overwhelming details. As we will flesh out in the last section \ref{debate} by referring to Miłkowski’s work \parencite{Mikowski2011, Mikowski2016}, a good computational explanation must link the formal story and the blueprint of the computing mechanism. Such articulation is not told in a formal theory of concurrent processes. 
Second, this first story of interactive systems restricts them to concurrent systems, which is only one dimension of interest when describing what current computers do. There are \textit{at least} two core dimensions left aside: what makes possible timing instructions and the connection between physical processes inside and outside the computing system.

\section{Reactive TMs: extending the original model}
\label{RTMS}

\subsection{Motivation}

More recently, a literature domain focused on a ``Reactive Turing machine'' has emerged 
\parencite{Andersen1997, Baeten2012, Baeten2013, Luttik2016, VanLeeuwen2006}. It reminds us that the purpose of Turing's \textit{a-machine} model was to propose a formal account of what is \textit{computable by effective means} (algorithmically computable). This formalization was achieved before the realization of the first digital computers. In a way reminiscent of Milner, the question is whether the TM model still fits computing practices decades later. The strategy chosen is to see whether \textit{extensions} of the original TM are sufficient to describe new computing practices and whether the obtained model is still equivalent to a TM. The strategy founds its frame within computability theory and reflects on its scope.
This literature domain that proposes extensions of the Turing machine to account for interactive computing systems may be traced back to seminal works on a ``Universal reactive machine'' \parencite{Andersen1997}. In that respect, although pointing at the specificity of interactional behaviour, the main framework still relates to Turing's. Baeten, Luttik, and van Tilburg \parencite{Baeten2013} are looking for a  model of interactive computation, extending the classical TM with a process-theoretical notion of interaction related to Milner's previous work. 
The strategy involves questioning the relationship between such extensions and the Church-Turing thesis. As a reminder, the Church-Turing thesis states that a computable function by effective means is computable by a Turing machine. The community interested in Reactive Turing machines asks the following question: can the Church-Turing thesis also be extended? Van Leeuwen and Wiedermann \parencite*{VanLeeuwen2001} focus on the possible extension of the Church-Turing thesis to account for interactive computing: ``We will motivate the need for a reconsideration of the classical Turing machine paradigm and formulate an extension of the Church-Turing thesis'' \parencite{VanLeeuwen2001}.

What is at stake is whether the Church-Turing thesis holds given warranted new models of computation: ``Is the Church-Turing thesis as we know it still applicable to the novel ways in which computers are now used in modern information technology? Will it hold for the emerging computing systems of the future?'' \parencite{VanLeeuwen2001}. 
The Church-Turing thesis originally did not entail a claim about computing in general (what computers do and will do) but only about \textit{effective computation}. Therefore, it does not follow that we should ask the Church-Turing thesis for answers on what computing is. Replacing a question about computing with a question about computation is the mark of a specific formal perspective within the frame of computability theory. Understanding \textit{computing} and its evolution from a formal perspective consist of questioning \textit{what can be computed} and seeing if there is another notion of computation than effective computation \textit{in the sense of Church-Turing}.

>>>>

\begin{artengenv2auth}{Alice Martin, Mathieu Magnaudet, Stéphane Conversy}
	{Modelling interactive computing systems: Do we have a~good theory of what computers are?}
	{Modelling interactive computing systems\ldots}
	{Modelling interactive computing systems: Do we have a~good theory\\of what computers are?}
	{\textsuperscript{1}Departments of Philosophy and Mathematics, Carnegie Mellon University\\
		\textsuperscript{2}Copernicus Center for Interdisciplinary Studies, Jagiellonian University}
	{Computers are increasingly interactive. They are no more transformational systems producing a~final output after a~finite execution. Instead, they continuously react in time to external events that modify the course of computing execution. While philosophers have been interested in conceptualizing computers for a~long time, they seem to have paid little attention to the specificities of interactive computing. We propose to tackle this issue by surveying the literature in theoretical computer science, where one can find explicit proposals for a~model of interactive computing. 
	In that field, the formal modelling of interactive computing systems has been brought down to whether the new interaction models are reducible to Turing Machines. There are three areas where interaction models are framed. The comparison between TMs and interactive system models is at stake in all of them. These areas are namely some works on concurrency by Milner, on Reactive Turing Machines, and on interaction as a~new computing paradigm. For each of the three identified models, we present its motivation, sum up its account for interaction and its legacy, and point out issues regarding the understanding of computers. The survey shows difficulties for epistemologists. The reason is that these analyses focus on the formal equivalence between interactive models of computation and classic ones. Such a~project is different from addressing how a~computing machine can be interactive: in other words, which mechanisms allow it.}
	{philosophy of computing, models of computation, interactive computing, computing mechanism, computational mechanistic explanation.}
	{%
		{\flushright\subbold{Alice Martin}\\\subsubsectit\small{ENAC, Université de Toulouse}\par}%
		{\flushright\subbold{Mathieu Magnaudet}\\\subsubsectit\small{ENAC, Université de Toulouse}\par}%
		{\flushright\subbold{Stéphane Conversy}\\\subsubsectit\small{ENAC, Université de Toulouse}\par}%
	}

%\documentclass[a4paper, 11pt, twoside]{article}
%
%%Citation-related commands
%%PP\usepackage{natbib}
%%PP\setcitestyle{authoryear,open={(},close={)}} 
%%PP \bibliographystyle{agsm}
%
%\usepackage[utf8]{inputenc}
%\usepackage{hyperref}
%%PP \renewcommand\harvardyearleft{\unskip, }
%%PP \renewcommand\harvardyearright[1]{.}
%%PP \hypersetup{hidelinks}
%
%\usepackage[top=3cm, bottom=3cm, left=3cm, right=3cm]{geometry}
%%\usepackage{amsmath}
%%\usepackage{amssymb}
%%\usepackage{mathrsfs}
%\usepackage{lmodern}
%%\usepackage[latin1]{inputenc}
%\usepackage[T1]{fontenc}
%\usepackage{dsfont}
%\usepackage{listings}
%\usepackage[table]{xcolor}
%%\usepackage{mathrsfs}
%\usepackage{float}
%\usepackage[justification = centering]{caption}
%%\usepackage{amsmath}
%\usepackage{graphicx}
%%\usepackage{lmodern}
%%\usepackage{bm} 
%
%%PP added
%\usepackage[style=bath, citestyle=authoryear-comp, backend=biber]{biblatex} %uwaga: wymaga zmiany procesora bibliografii na biber
%\addbibresource{Martin.bib}
%%end of PP added
%
%
%\title{Modelling Interactive Computing Systems: Do We Have a~Good Theory of What Computers Are?}
%%\date{}
%\author{Alice Martin, Mathieu Magnaudet, Stéphane Conversy}
%
%\begin{document}
%
%\maketitle
%
%Computers are increasingly interactive. They are no more transformational systems producing a~final output after a~finite execution. Instead, they continuously react in time to external events that modify the course of computing execution. While philosophers have been interested in conceptualizing computers for a~long time, they seem to have paid little attention to the specificities of interactive computing. We propose to tackle this issue by surveying the literature in theoretical computer science, where one can find explicit proposals for a~model of interactive computing. 
%In that field, the formal modelling of interactive computing systems has been brought down to whether the new interaction models are reducible to Turing Machines. There are three areas where interaction models are framed. The comparison between TMs and interactive system models is at stake in all of them. These areas are namely some works on concurrency by Milner, on Reactive Turing Machines, and on interaction as a~new computing paradigm. For each of the three identified models, we present its motivation, sum up its account for interaction and its legacy, and point out issues regarding the understanding of computers. The survey shows difficulties for epistemologists. The reason is that these analyses focus on the formal equivalence between interactive models of computation and classic ones. Such a~project is different from addressing how a~computing machine can be interactive: in other words, which mechanisms allow it. \\
%
%
%Keywords: philosophy of computing, models of computation, interactive computing, computing mechanism, computational mechanistic explanation

@@

In the philosophy of computing, we are paying increased attention to a~set of new features of computers. This set has led to the introduction of a~new label for these computing machines: they are referred to as interactive computing machines \parencite{Dodig-Crnkovic2011, Goldin2006, Soare2013, VanLeeuwen2001, Wegner1997}. The set of new features can be captured in the following statement made in a~2011 paper by Gordana Dodig-Crnkovic (our highlights):

\myquote{
 Present day computers are very different from the early stand-alone calculators designed for mechanizing mathematical operations. They are largely \emph{used for communication in world-wide networks} and variety of information processing and knowledge management. Moreover, they play \emph{an important role in the control of physical processes and thus connect to the physical world}, especially in automation and robotics. [\ldots] Computational processes are nowadays \emph{distributed, reactive, agent-based, and concurrent}. The main criterion of success of the computation is not its termination, but its \emph{response to the outside world}, its speed, generality and flexibility; adaptability, and tolerance to noise, error, faults, and damage. \parencite{Dodig-Crnkovic2011}
 }

Historically, the concept of interaction was introduced by a~computer scientist, Milner, in the 1970s-1980s \parencite{Milner1975, Milner1982, Milner1993, Milner1999}. At first, an \emph{interactive computing system} was defined as a~system where several threads execute instructions in parallel while being able to synchronize and communicate at certain moments of the execution. Since then, the characteristics of computer systems have continued to evolve, and by ``interactive'' we refer today to a~broader set of properties that can be grouped as follows: the ability to continuously react in time to external events that modify computing execution. 
This class of computers deserves all our attention since they are ubiquitous. Every computer system today is designed to respond to external events in a~predictable way and according to temporal constraints. In any case, what distinguishes this class of so-called interactive computing machines from the classical computer systems that preceded them is that they are no longer purely \emph{transformational systems}. A~\emph{transformational system} is a~classical computing device that, given a~set of inputs, produces a~final output after a~finite execution. 
This evolution of computing complicates the answer to \textit{what a~computer is}. The question is well-known in the philosophy of computing \parencite{Piccinini2008, Rapaport2018, Smith2002}. As already noted, many answers to the question distort it and answer the question of \textit{what a~computation is}, immediately projecting the field of investigation into the theory of computability:

\myquote{
A~fairly obvious, trivial, and almost-circular definition of ‘computer’ says that a~computer is a~machine that computes. The natural next question is: What does it mean to compute? But this shifts the burden of answering our question away from what computers are to the topic of what computation is. Many of the objections to various theories about computers are really objections to what counts as a~computation. \parencite{Rapaport2018}. 
}

This leaves us with the specific issue we want to address. We ask whether models of computation for interaction allow us to answer the question of what an actual (necessarily interactive) computer is. Current computers come in various forms and we chose in this paper to restrict our concerns to a~delimited notion of \textit{interaction}, as defined in Human-Computer Interaction \parencite{Myers1994, Dearden1997, Beaudouin-Lafon2006,  Hornbaek2017, Basman2018}, and target a~specific set of ubiquitous computing devices---those interacting with humans, e.g., through digital interfaces. We will not elaborate on \textit{analog computing} \parencite{Bielecki2019} and \textit{natural computation} \parencite{Dodig-Crnkovic2011, MacLennan2003}.

To tackle the issue of interactive computing devices, we propose here an approach that, to the best of our knowledge, has not been proposed so far: we want to examine the models of computation proposed in theoretical computer science to think about interactive computing systems. We offer a~literature survey where one can find \textit{explicit theories of interaction} \footnote{We insist on our two criteria: \textit{explicit theories of interaction} in \textit{theoretical computer science}. We have in mind the fact that other communities e.g., the engineering community on reactive systems, are related to our topic but they have not conceptualized \textit{interaction} as such.}. We show that the formal modelling of interactive computing systems has been brought down to whether the new interaction models are reducible to Turing’s \emph{a-machines} \parencite{Turing1937}---we will refer to them as Turing Machines (TMs). 
Questioning the theoretical bounds of the Turing Machine in computer science when faced with the existence of interactive computing devices has been explored at least since Milner's work on communicating and mobile systems \parencite{Milner1993, Milner1999}. To the best of our knowledge, there are three areas where interaction models are framed as such. These areas are some works (i) on concurrency by Milner and his followers \parencite{Milner1999, Milner2006}, (ii) on Reactive Turing Machines \parencite{Andersen1997, Baeten2013, VanLeeuwen2001, VanLeeuwen2006}, and (iii) on interaction as a~new non-algorithmic computing paradigm \parencite{Goldin2006, Wegner1997, Wegner2003}. For each of the three identified models, we:

@@

We then want to show how these approaches, which belong to a~formal approach, cannot provide an answer to the question of what computers are, and for two reasons. On the one hand, these models of computation have focused their attention on whether interactive models are reducible to models of classical computation---par excellence, the Turing machine. Proving (or not) that an interactive property can be formalized as a~computable property in the classical Turing’s sense does not answer the question of how an interactive property comes into existence and can be the object of execution. On the other hand, and this is a~correlate, these models do not propose a~basis for a~mechanistic explanation of the very possibility of an interactive computing system. With only formal models of interactive computation, we might run the risk of not offering an adequate conceptualization of current computers. Therefore, we end up proposing to take the distinction seriously between \emph{models of computation} and \emph{mechanistic computational explanations}, as presented by Miłkowski \parencite*{Mikowski2011, Mikowski2014}.

\section{Milner introduces a~distinction between interactional and computational behavior}
\label{Milner}
%\vskip -2\baselineskip
\subsection{Motivation}

Milner was the one who introduced the concept of interaction in computer science. He summarized his motivations in a~famous Turing Award speech \parencite{Milner1993}. Milner was concerned with the logical foundations of computing inherited from Turing. He was preoccupied with the idea that computing practices had evolved since the birth of computing, notably in terms of architecture. He took seriously the possibility that the logical foundations dating back to the thirties may not match the growing challenges of his time and may require additional concepts.

Milner \parencite*{Milner2006} pointed out  that the logical foundations of computing offered by Turing \parencite*{Turing1937} were previous to the first physical computers and that computer science is grounded in \emph{logic} and \emph{engineering}. On the engineering side, computer science was inherited from von Neumann’s pioneering work \parencite{Godfrey1993, Aspray1990}. Only one thing could happen at once in an early von Neumann's computer. Nevertheless, there was more to computing than von Neumann's architecture \parencite{Backus1978, Milner2006}. A~growing interest in dealing with concurrency in the sixties and seventies made sequential programming less warranted. Therefore, to Milner, the logical foundations of computing were to evolve. The main flaw of the early logical foundations was the reduction of computing processes to the concept of an algorithm, which tends to associate computing with mere calculation without taking concurrent activity into account. Because of the evolution of computing engineering practice, Milner questioned whether the logical grounds of computing should evolve as well. Milner's thesis can be put in a~nutshell: ``this logical foundation has changed a~lot since Turing but harks back to him. To be more precise: (i) Computing has grown into informatics---the science of interactive systems; (ii) Thesis: Turing's logical computing machines are matched by a~logic of interaction'' \parencite{Milner2006}. Consequently, a~theory and new language to express concurrent activity were required: ``we must find an elementary model which does for interaction what Turing's logical machines do for computation'' \parencite{Milner2006}.

The need to define a~new computing theory is first displayed through the evolution of computing practice.
To sum up, Milner’s motivation and focus were the solving of concurrency issues in distributed systems, with the idea that the evolution of computing practices required new formal tools: ``Through the 1970s, I~became convinced that a~theory of concurrency and interaction requires a~new conceptual framework, not just a~refinement of what we find natural for sequential [algorithmic] computing'' \parencite{Milner1993}.

@@

Milner introduced the opposition between \textit{interactional} and \textit{computational} behaviour. Introducing the concept of interaction, Milner \parencite*{Milner1975, Milner1982, Milner1983} referred to concurrent message passing between agents. Milner's work coincided with Petri’s \parencite*{Petri1980} new model of concurrent processes, which generally intended to describe concurrency in information systems.\footnote{Concurrency theory emerged from Dick Karp's early work in the 1960s, grew with \parencite{Petri1980} and later work on transition systems \parencite{Nielsen1981, Glabbeek2004},
and has now developed into a~mature theory of reactive systems \parencite{Harel1985} with diverse network models \parencites[for an overview, see][]{Lee1998}{Lee2006_modelsofcomputation}}
 To Milner, interaction is more \textit{expressive} than a~TM, but it still describes an \textit{effective} procedure. Milner did not assert \textit{equivalence} between an interactive model and a~TM, but he introduced the topic \parencite{Milner1999} and left it unanswered. 
Four main differences between old (computational) and new (interactional) computing are made striking by Milner. First, in Milner’s words, a~Turing Machine prescribes a~behaviour to be executed. In contrast, new computing requires the description of an information flow between several system components. Second, old computing is characterized by a~\textit{hierarchical design} when current practice involves \textit{heterarchical} phenomena in the computing system. Third, in new computing, the designer cannot predict when agents will be triggered or the overall behaviour of the computing system. Fourth, the user is not merely looking for an end result in new computing practice. There is more than a~mathematical function to evaluate, as it used to be in old computing. The user instead interacts with the system, and the look for an end result is replaced by continuing interaction. Having taken stock of the evolution of computing practice on the engineering side, Milner examines its consequences on the logic foundations of computing. The pi-calculus and his work on the equivalence with automata, known as bisimulation, achieved this reflection on interactive processes \parencite{Milner1993, Milner1999} with a~formalism.

@@

Milner's work on interaction has become a~founding block in automata theory and concurrency theory. It installed the notion of a~transition system as the prime mathematical model to represent discrete behavior \parencite{Nielsen1981, Baldan2001, Glabbeek2004, Arbach2015}. It also showed that language equivalence was not the correct notion when comparing automata for interactive systems.
Instead, it should be replaced by a~notion of behavioral equivalence or bisimilarity~\parencite{Milner1999}. The pi-calculus has inspired research to derive a~language from it. The \textit{Pict} \parencite{Pierce2000} programming language is an example.

Milner's work is foundational and served as a~reference for anyone after him, reflecting on the need for a~new framework dedicated to new emerging computing practices. Milner insists on an essential reminder that we would like to consider. When modelling, the engineering practice matters and is to be articulated with the logical foundations of the model, should it involve elaborating a~new framework. Famously, Wegner and Goldin acknowledge that Milner was the first to introduce the idea that classic models of computation were insufficient. They argue that Milner did not state clearly whether computation in concurrent communicating systems (CCS) and the pi-calculus were reducible to Turing machines and algorithms \parencite{Wegner2003}. If one goes and looks at Milner’s Turing Award Speech, it seems true that classical computation translates into an interactive calculus. However, it is not stated whether any formula in the pi-calculus can be expressed in a~classical calculus like the lambda-calculus.

@@

Given the account of current computers that we are looking for, we see two limits in the lessons drawn from Milner.
First, we are looking for an explanation of the interactive computing phenomena at stake in a~computer. Therefore, the relation between layers of abstraction, from the computational to the physical, is crucial. However, to Milner, the physical layer of the machine is not of much interest, and the calculus of CCS needs to abstract away from the physical. As Milner puts it, informatics is about virtual symbols: ``physical systems tend to have permanent physical links; they have fixed structure. But most systems in the informatic world are not physical; their links may be virtual or symbolic'' \parencite{Milner1999}. From our perspective, abstracting away from the physical world comes at some cost for an explanation. A~complete explanation of a~computing system can hardly be provided in details within a~single understandable abstraction, since a~computing sytems is extremely  multi-layered~\parencite{Nisan2005, Plato2018}. Therefore, an explanation of a~computing system is necessarily a~trade-off between understandability and overwhelming details. As we will flesh out in the last section \ref{debate} by referring to Miłkowski’s work \parencite{Mikowski2011, Mikowski2016}, a~good computational explanation must link the formal story and the blueprint of the computing mechanism. Such articulation is not told in a~formal theory of concurrent processes. 
Second, this first story of interactive systems restricts them to concurrent systems, which is only one dimension of interest when describing what current computers do. There are \textit{at least} two core dimensions left aside: what makes possible timing instructions and the connection between physical processes inside and outside the computing system.

\section{Reactive TMs: extending the original model}
\label{RTMS}
%\vskip -2\baselineskip
\subsection{Motivation}

More recently, a~literature domain focused on a~``Reactive Turing machine'' has emerged 
\parencite{Andersen1997, Baeten2012, Baeten2013, Luttik2016, VanLeeuwen2006}. It reminds us that the purpose of Turing's \textit{a-machine} model was to propose a~formal account of what is \textit{computable by effective means} (algorithmically computable). This formalization was achieved before the realization of the first digital computers. In a~way reminiscent of Milner, the question is whether the TM model still fits computing practices decades later. The strategy chosen is to see whether \textit{extensions} of the original TM are sufficient to describe new computing practices and whether the obtained model is still equivalent to a~TM. The strategy founds its frame within computability theory and reflects on its scope.
This literature domain that proposes extensions of the Turing machine to account for interactive computing systems may be traced back to seminal works on a~``Universal reactive machine'' \parencite{Andersen1997}. In that respect, although pointing at the specificity of interactional behaviour, the main framework still relates to Turing's. Baeten, Luttik, and van Tilburg \parencite{Baeten2013} are looking for a~ model of interactive computation, extending the classical TM with a~process-theoretical notion of interaction related to Milner's previous work. 
The strategy involves questioning the relationship between such extensions and the Church-Turing thesis. As a~reminder, the Church-Turing thesis states that a~computable function by effective means is computable by a~Turing machine. The community interested in Reactive Turing machines asks the following question: can the Church-Turing thesis also be extended? Van Leeuwen and Wiedermann \parencite*{VanLeeuwen2001} focus on the possible extension of the Church-Turing thesis to account for interactive computing: ``We will motivate the need for a~reconsideration of the classical Turing machine paradigm and formulate an extension of the Church-Turing thesis'' \parencite{VanLeeuwen2001}.

What is at stake is whether the Church-Turing thesis holds given warranted new models of computation: ``Is the Church-Turing thesis as we know it still applicable to the novel ways in which computers are now used in modern information technology? Will it hold for the emerging computing systems of the future?'' \parencite{VanLeeuwen2001}. 
The Church-Turing thesis originally did not entail a~claim about computing in general (what computers do and will do) but only about \textit{effective computation}. Therefore, it does not follow that we should ask the Church-Turing thesis for answers on what computing is. Replacing a~question about computing with a~question about computation is the mark of a~specific formal perspective within the frame of computability theory. Understanding \textit{computing} and its evolution from a~formal perspective consist of questioning \textit{what can be computed} and seeing if there is another notion of computation than effective computation \textit{in the sense of Church-Turing}.
